# Model Error

## Systematic Error and Random Error

Assume $\mathbf{X}$ is $n \times p$ observation matrix and $\mathbf{y}$ is response variable, we have:

$$\mathbf{y}=f(\mathbf{X})+\mathbf{\epsilon}$$

where $\mathbf{\epsilon}$ is the random error with a mean of zero. The function $f(\cdot)$ is our modeling target, which represents the information of Y that can be explained by X. The main goals of estimating $f(\cdot)$ are inference and prediction, or sometimes both. In general, there is a trade-off between flexibility and interpretability of the model. So data scientists need to comprehend the delicate balance between the two. 

Depending on the modeling purposes, the requirement for interpretability varies. If the prediction is the only goal, then as long as the prediction is accurate enough,  the interpretability is not under consideration. In this case, people often use "black box" model, such as random forest, boosting tree, neural network, support vector machine and so on. These models are very flexible but nearly impossible to explain. Their predictive accuracy is usually higher on the training set, but not necessary when it predicts. It is not surprising since those models have a huge number of parameters and high flexibility that they can memorize the entire training data. A paper by Chiyuan Zhang et. al. in 2017 pointed out that "Deep neural networks (even just two-layer net) easily fit random labels" (https://arxiv.org/abs/1611.03530). The traditional forms of regularization, such as weight decay, dropout, and data augmentation, fail to control generalization error. It poses a conceptual challenge to statistical theory and also calls our attention when we use such black-box models.

Assume we have $\hat{f}$ which is an estimator of $f$. Then we can further get  $\mathbf{\hat{y}}=\hat(f)(\mathbf(X))$. The predicted error is divided into two parts, systematic error and random error:

$$
E(\mathbf{y}-\hat{\mathbf{y}})^{2}=E[f(\mathbf{X})+\mathbf{\epsilon}-\hat{f}(\mathbf{X})]^{2}=\underset{\text{(1)}}{\underbrace{E[f(\mathbf{X})-\hat{f}(\mathbf{X})]^{2}}}+\underset{\text{(2)}}{\underbrace{Var(\mathbf{\epsilon})}}
  \label{eq:error}$$
  

In the above equation, (1) is the systematic error, It exists because $\hat{f}$  usually does not completely model the "systematic relation" between X and y, where the system relation refers to the stable relationship that exists on different samples. This part of the error can be improved by improving the model; (2) is the random error which represents the part of the response that can not be explained by current input data. A more complex model does not reduce the error. The biggest problem with black-box models is to fit random error as well, i.e., over-fitting. The notable feature of random error is that it can not be repeated on different samples. So a way to determine whether there is overfitting is to reserve a part of the sample as a test set and then check the performance of the trained model on the test data. Note that overfitting is a general problem, and any model may be overly fitted. Because black-box model usually has a large number of parameters,  it is more suspectable to over-fitting. 

The systematic error can be further decomposed as:

\[
\begin{array}{ccc}
E[f(\mathbf{X})-\hat{f}(\mathbf{X})]^{2} & = & E\left(f(\mathbf{X})-E[\hat{f}(\mathbf{X})]+E[\hat{f}(\mathbf{X})]-\hat{f}(\mathbf{X})\right)^{2}\\
 & = & E\left(E[\hat{f}(\mathbf{X})]-f(\mathbf{X})\right)^{2}+E\left(\hat{f}(\mathbf{X})-E[\hat{f}(\mathbf{X})]\right)^{2}\\
 & = & [Bias(\hat{f}(\mathbf{X}))]^{2}+Var(\hat{f}(\mathbf{X}))
\end{array}
\]


The systematic error consists of two parts,  $ Bias(\hat{f}(\mathbf{X}))$ and  $Var (\hat{f} (\mathbf{X}))$. To minimize the systematic error, we need to minimize the bias and variance. The bias represents the error caused by the approximation of the reality of the model, which may be very complex. For example, linear regression assumes that there is a linear relationship between the independent variable and the response, but the perfect linear relationship in real life is not common. The relationship between x and fx in the following figure is non-linear. Therefore, despite a large sample size, linear regression can not give the accurate prediction. In other words, in this case, the prediction of the linear regression model has a high bias.

```{r}
library(grid)
library(lattice)
library(ggplot2)
source("https://raw.githubusercontent.com/happyrabbit/DataScientistR/master/R_Code/multiplot.r")
# randomly simulate some non-linear samples
x=seq(1,10,0.01)*pi
e=rnorm(length(x),mean=0,sd=0.2)
fx<-sin(x)+e+sqrt(x)
dat=data.frame(x,fx)
# plot fitting result
ggplot(dat,aes(x,fx))+ 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

The estimated parameters will be different for the different training data. Intuitively, the estimated variance indicates that if we fit the same model with different data sets, how much the estimated results will change. Ideally, the change is small. For high variance models, small changes in the training data result in very different estimates. In general, a model with high flexibility also has high variance., such as the CART tree, and the initial boosting method. The Random Forest and Gradient Boosting Model aim to reduce the variance by summarizing the results obtained on different samples. 

The blue curve in the figure below is obtained by fitting the above nonlinear observations by a smoothing method, which is highly flexible and can fit the current data tightly:

```{r}
ggplot(dat,aes(x,fx))+geom_smooth(span = 0.03)
```

However, this method has a high variance, and if we simulate different subsets of the sample, the result curve will change significantly:

```{r, warning=FALSE}
# set random seed
set.seed(2016)
# sample part of the data to fit model
# sample 1
idx1=sample(1:length(x),100)
dat1=data.frame(x1=x[idx1],fx1=fx[idx1])
p1=ggplot(dat1,aes(x1,fx1))+geom_smooth(span = 0.03)
# sample 2
idx2=sample(1:length(x),100)
dat2=data.frame(x2=x[idx2],fx2=fx[idx2])
p2=ggplot(dat2,aes(x2,fx2))+geom_smooth(span = 0.03)
# sample 3
idx3=sample(1:length(x),100)
dat3=data.frame(x3=x[idx3],fx3=fx[idx3])
p3=ggplot(dat3,aes(x3,fx3))+geom_smooth(span = 0.03)
# sample 4
idx4=sample(1:length(x),100)
dat4=data.frame(x4=x[idx4],fx4=fx[idx4])
p4=ggplot(dat4,aes(x4,fx4))+geom_smooth(span = 0.03)
multiplot(p1,p2,p3,p4,cols=2)
```

Fitting the linear model using the same four subsets, the result barely changes:

```{r,warning=FALSE}
p1=ggplot(dat1,aes(x1,fx1))+ 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
p2=ggplot(dat2,aes(x2,fx2))+ 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
p3=ggplot(dat3,aes(x3,fx3))+ 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
p4=ggplot(dat4,aes(x4,fx4))+ 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
multiplot(p1,p2,p3,p4,cols=2)
```